---
title: "R Programming Basics - part 2"
output:
  pdf_document: null
  html_document:
    toc: yes
  pdf_presentation: null
  slidy_presentation:
    css: slidy.css
    widescreen: yes
  word_document: null
---

## Outline

- Flow Control
- Functions
- R for statistics

# Flow Control

## Conditional execution

### *if* statement 

If a logical condition holds, execute one or more statements

syntax: if (condition) {executable statements}

```{r eval=T}
nclasses=2
fulltime=FALSE
if (nclasses >= 3){
  fulltime=TRUE}
fulltime
```

### *if else* pair

If a condition holds, execute one set of statements, or else
execute a second set of statements.

syntax: if (condition) {
             executable statements 1
	     } else {
	     executable statements 2
	     }

```{r eval=T}
fulltime=NA
nclasses=1
if (nclasses >= 3) {fulltime=TRUE
} else {fulltime=FALSE}
fulltime
```

It is essential to begin the second line of this if/else pair as
 "} else".  In this case the R interpreter "waits" for the "}", and
 sees it paired with the "else".
  The "if" statement would be completed if the "}" was 
 included on the end of the first line.
 
Often one wants to use several "else if" clauses.  The following
example takes a numerical grade, and assigns a letter grade according
to the Dalhousie common grade scale.

```{r eval=T}
  grade=91.5
  Lgrade=NA
if (grade >=90) {Lgrade="A+"
                 print("Good job!")
} else 
    if (grade >= 85) {Lgrade="A"
} else 
    if (grade >= 80) {Lgrade="A-"
} else 
    if (grade >= 77) {Lgrade="B+"
} else 
    if (grade >= 73) {Lgrade="B"
} else 
    if (grade >= 70) {Lgrade="B-"
} else 
    if (grade >= 65) {Lgrade="C+"
} else 
    if (grade >= 60) {Lgrade="C"
} else 
    if (grade >= 55) {Lgrade="C-"
} else 
    if (grade >= 50) {Lgrade="D"
    } else {Lgrade="F"; print("Too bad!")}
Lgrade
```
Example: determine if a number is positive, negative or 0.

```{r eval=T}
a <- 0
if (a < 0) {
   print("a is a negative number")
} else if (a > 0) {
   print("a is a positive number")
} else {
   print("a is zero")
}
```



### ifelse

#### *ifelse* is a compact version of *if ... else ...*
#### Syntax
```{r eval = F}
ifelse(condition, statement1, statement2)
```

There is a subtle difference in that *ifelse* works on vectors.

#### Examples

```{r}
#find the minimum of scalars a and b
a=rnorm(1); a
b=rnorm(1); b
minab=ifelse(a<b,a,b); print(minab)

#find the component by component minimum of vectors a and b
a=rnorm(10); 
b=rnorm(10); 
vecminab=ifelse(a<b,a,b); print(cbind(a,b,vecminab))

#determine if the elements of a vector are evenly divisible by 2
vec = 1:12
evenOdd = ifelse(vec %% 2 == 0, "even", "odd")
print(evenOdd)

```

## Logical Operators

& (and), | (or), == (equals), ! (not)

- get help on logical operators
```{r}
#?"=="
```

## Truth tables

Suppose that $x$, $y$ are logical variables, taking values
$T$ or $F$.  The following "truth table" gives the values of
the functions $x\&y$ (x and y), $x|y$ (x or y), and $!x$ (not x).

```{r echo=F}
x=c(T,T,F,F)
y=c(T,F,T,F)
xANDy=c(T,F,F,F)
xORy=c(T,T,T,F)
notx=c(F,F,T,T)
M=cbind(x,y,xANDy,xORy,notx)
dimnames(M)[[2]]=c("x","y","x&y","x|y","!x")
M
```
These basic functions can be built up into more complex expressions.

Example: $(x|y)\&z$

```{r echo=F}
xory=c(T,T,F,F)
z=c(T,F,T,F)
l1=c(T,F,F,F)
M=cbind(xory,z,l1)
dimnames(M)[[2]]=c("x|y","z","(x|y)&z")
M
```

Example:$(x\&y)|z$

```{r echo=F}
xandy=c(T,T,F,F)
z=c(T,F,T,F)
l2=c(T,T,T,F)
M=cbind(xandy,z,l2)
dimnames(M)[[2]]=c("x&y","z","(x&y)|z")
M
```

It is often useful to include brackets in order to specify the logical expression that you want.
For example, suppose that you are interested to know whether (one or the other of x and y are TRUE)
and (z is TRUE).  Then you want to evaluate (x|y)&z.  Suppose that you forget to include the
brackets, and instead evaluate x|y&z.  Are these the same?  Sometimes yes, sometimes no.

```{r}
x=c(rep(T,4),rep(F,4))
y=rep(c(rep(T,2),rep(F,2)),2)
z=rep(c(T,F),4)
cbind(x,y,z,(x|y)&z,x|y&z,x|(y&z))
```


### Using logical values to subset data

- Examples:

```{r}
vec = c(1, 3, 6, 2, 5, 10, 11, 9)

#return all values which are evenly divisible by 3
vec[vec%%3 == 0]

#return all values which are not evenly divisible by 3
vec[!vec%%3 == 0]
vec[!((vec%%3) == 0)]  #better to use brackets if you're unsure of operator precedence

# return all the values in vec which are greater than 5
vec[vec > 5]

# use "cars" data set, return all the rows where dist is less than 20
cars[cars$dist < 20, ]

# return all the rows where speed is 10
cars[cars[,1] == 10, ]
```

###Example: is a specified year a leap year? (Uses &, |, and !)

```{r}
year = 2018
if ( (year %% 4 == 0 & year %% 100 != 0) | year %% 400 ==0){
    print(paste(year,"is a leap year"))
  } else {
    print("no")
  }
```

Here is essentially the same thing, where the logical operations are more clearly identified.

```{r}
  year=2018
  x=year %% 4 == 0
  y=year  %% 100 != 0
  z=year %% 400 ==0
  condition=x&y|z
  c(x,y,z,condition)
  if(condition) print(paste(year,"is a leap year"))
  ```




####Example:  find records in a database
```{r}
rm(list=ls()) #remove all objects
ls()
subjectno=c(1:8)
#enter some data for first and last name, age
firstname=c("Dick","Jane","","Jian","jing","Li","John","Li")
lastname=c("Tracy","Doe","Smith","Yuan","Xian","Li","Doe","")
age=sample(c(18:35),8) #assign random ages from 18 through 35
data=data.frame(subject=subjectno,firstname=firstname,
      surname=lastname,age=age)
rm("subjectno","firstname","lastname","age")
ls()
attach(data) #attach the dataframe data
data
#find subjects whose surname is "Li"
subset1=data[surname=="Li",];subset1
#Is there a Jane Doe in the database?
subset2=data[surname=="Doe"&firstname=="Jane",];subset2
#find subjects whose given or surname is missing ("")
subset3=data[surname=="" | firstname=="",];subset3
#Find the subjects who are older than 29 years?
data[age>29,]
#find all subjects whose first name starts with "J"
#use substr(chvector,1,1) to extract the first character
#of each element of the character vector chvector
subset4=data[substr(firstname,1,1)=="J",];subset4
#find all subjects whose first name starts with either "J" or "j"
subset5=data[substr(firstname,1,1)=="J"|substr(firstname,1,1)=="j",];subset5
detach(data) #detach the dataframe data
```

# Flow control 

  *for*, *while*, *repeat*, *break* and *next*

  *for*, *while* and *repeat* are iterative constructs, meaning that a collection of
   R statements are repeated.
  
### *for* is the most common loop structure
-Syntax
```{r eval=F}
for (var in range) {
    statements
}
```


- Examples:

```{r}
# sum the numbers 1:k
k=12
mysum=0  #initialize the sum to 0
for (i in 1:k) {
  mysum=mysum+i}
print(mysum)
```

- Example: Find which of the years 2000 through 2020 are leap years
```{r }
for (i in 2000:2020){
  if ( (i %% 4 == 0 & i %% 100 != 0) | i %% 400 ==0){
    print(paste(i,"is a leap year"))
  }
}

```

Here is essentially the same thing, where the logical operations are more clearly

```{r}
for (i in 2000:2020){
  x=i %% 4 == 0
  y=i %% 100 != 0
  z=i %% 400 ==0
  condition=x&y|z
  print(c(x,y,z,condition))
    if(condition) print(paste(i,"is a leap year"))

}
```


 *Example*: Calculate the  *inner product* of two vectors.
Suppose that $u$ and $v$ are each numeric vectors of length $n$.
Then the inner product of $u$ and $v$ is defined as

$$\sum_{i=1}^n u_i v_i$$

That is, we sum the products of the corresponding elements of the
two vectors.  This is easy to program using a *for* loop.

```{r}
u=1:10
v=-10:-1
u*v

prod=0
for (i in 1:10)prod=prod+u[i]*v[i]  #note no {} needed if everying on same line
print(prod)

#verify
sum(u*v)
```

Example: *matrix multiplication*
Suppose that $U$ is a matrix with $m$ rows and $n$ columns, and that
$V$ is a matrix with $n$ rows and $p$ columns.  The the matrix product
$P=UV$ is a matrix with $m$ rows and $p$ columns, and the entry in the
$i$'th row and $j$'th column of $P$ is

$$P_{ij} = \sum_{k=1}^n U_{ik} V_{kj}$$

Define some conforming matrices U and V, in this example, U being 4 by 3 and V being 3 by 3
The product $P$ is a 4 by 3 matrix.  Find $P_{2,3}$, the element in the 2nd row and 3rd column of $P$.
By definition 

$$P_{2,3} = \sum_{k=1}^n U_{2k} V_{k3}$$

```{r}
U=matrix(1:12,byrow=T,ncol=3)
U # a 4 by 3 matrix
V=matrix(c(rep(1,3),c(-1,0,1),-3:-1),byrow=T,ncol=3)
V # a 3 by 3 matrix
U2=U[2,] #second row of U
V3=V[,3] #third column of V
P23=0
for (k in 1:3) P23=P23+U2[k]*V3[k]
P23  
```

We want to calculate $P_{i,j}$ for each pair of indices (i,j).
Essentially

```{r eval=F}
 for (i in 1:4){
   #for each i in (1,2,3,4), do something
  for (j in 1:3){
    #for each j in (1,2,3) do something
     #calculate Pij here
  } # end of j for loop
    } # end of i for loop
```

This can be calculated in R using three "nested" for loops.

```{r}
#result will be a 4 by 3 matrix, having 12 elements
P=matrix(rep(0,12),byrow=T,ncol=3) #initialize product with 0's
for (i in 1:4){ #for each i, execute code until the closing }
  for (j in 1:3){ #for each j, execute code until the closing }
    for (k in 1:3){ # for each k, execute code until the closing }
      P[i,j]=P[i,j]+U[i,k]*V[k,j]
      }}}
P

#verify using the builtin R matrix multiplication operator "%*%".

U%*%V

```
 Executing loops, in this case, "for" loops,  in R is very inefficent as compared to many computer languages, such as Java or C, and using the built in functions (in this case %*%) is always recommended.  That said, you need to know how to use loops effectively, in order to be able to do things in addition to calling the built in procedures. 


The for loop in R is more general than the simlar construct in most
programming languages.  The general structure is 

```{r eval=F}
for (var in set) {
    statements
}
```
 

Where "set" is an arbitrary set, and the instructions are
evaluated for each element of the set.  

```{r}
set=c("this","that","cat","mouse","male","female")
for (var in set) {
   print(var)
}
```

#### The basic definitions and a few simple examples are given below for the *while*, *repeat*, *break*, *next*
#### statements/constructs.

More meaningful examples will be given in the material on user defined functions.



#### *while*

#### Syntax
```{r eval=F}
while (condition){
  statement
}

```

#### Example
```{r}
x = 5
while(x <= 20){
  print(x)
  x = x+5
}

```

#### *repeat*

#### Syntax: repeat {statement}
#### need to use **break**
```{r}
x = 5
repeat{
  print(x)
  x = x+5
  if (x > 20) break
}

```

#### "break" and "next"
#### "break": break the current loop
```{r}
for (i in 1:6){
  if (i==5){
    break
  }
  print(i)
}
```

#### "next": skip to next iteration
```{r}
for (i in 1:6){
  if (i==5){
    next
  }
  print(i)
}

```


# User-defined functions
-Syntax
```{r eval=F}
funtionname= function(arg1,arg2,...){
  statements
  return(something)
}

```

-Examples:

```{r}
SumSquare = function(x,y){
    val = x^2+y^2
    return(val)
}
SumSquare(3,4)
```

```{r}
SumSquares = function(x){
# function to create the sum of squares of elements of x
    temp=0
    for (i in 1:length(x)){
#    print(c(i,x[i],x[i]^2))
    temp=temp+x[i]^2}
    return(temp)
}

data=c(4,1,-2,5)
SumSquares(data)

```{r}
SumSquares1 = function(x){
# another way to do the same thing
    temp=0
    for (i in x){
    print(c(i,i^2))
    temp=temp+i^2}
    return(temp)
}

SumSquares1(data)

SumSquares2 = function(x){
# if statement checks that the input argument x
# is a numeric vector.  If it is not, print an
# error message, and return a NULL value
    if(!is.vector(x)|!is.numeric(x)){
       print("x should be a numeric vector")
       return(NULL)}
# otherwise, return the sum of the squared elements of x
    temp=0   
    for (i in 1:length(x))temp=temp+x[i]^2
    return(temp)
}

SumSquares2(data)
SumSquares2(c("a","b"))
SumSquares2(matrix(1:4,byrow=T,ncol=2))
```



# Exercise: create a function for finding leap years
- input: startYear, endYear
- output: return a vector of all the leap years between startYear and endYear
- Using a for loop

```{r}
leapYears1 = function(startYear,endYear){
  output = NULL
#uses a for loop
  for (year in c(startYear:endYear)){
  if ( (year %% 4 == 0 & year %% 100 != 0) | year %% 400 ==0){
    output = c(output,year)
  }}
  return(output)
}

leapYears1(2018,2028)
```


- Use *while* or *repeat* for looping

```{r}
leapYears2 = function(startYear,endYear){
  results = c()
#uses a while loop, as opposed to a for loop
  year = startYear
  while (year <= endYear){
  if ( (year %% 4 == 0 & year %% 100 != 0) | year %% 400 ==0){
    results = c(results,year)
  }
    year = year + 1
  }
  return(results)
}

leapYears2(2018,2028)

```

```{r}
leapYears3 = function(years=2019){
  #uses built in vectorized logical indexing
  return(years[(years %% 4 == 0 & years %% 100 != 0) | years %% 400 ==0])}

leapYears3(2018:2028)
test=leapYears3(1867:2019)
test
```
-Example: function calling another function

Recall that the formula for the sample variance of $x_1, x_2, \ldots , x_n$ is

$$s^2 =\frac{1}{n-1} \sum_{i=1}^n (x_i-\bar x)^2$$

where $\bar x = \sum_{i=1}^n x_i / n$.

```{r}
mymean=function(x){
#returns the sample mean of the values in the vector x
  mysum=0
  for (i in x){
    mysum=mysum+i
  }
  return(mysum/length(x))
}

myvar=function(x){
  n=length(x)
# subtract the mean from the values in x
  data=x-mymean(x)
# sum the squares of the entries of data
  myvar=SumSquares(data)
# divide the sum of squares by n-1  
  myvar=myvar/(n-1)
  return(myvar)
  }

myvar(data)

#check using the built in function var

var(1:7)
```

-How fast is our user defined function as compared to the builtin function *var*?

```{r}
v=rnorm(50000) #vector of 50000 observations from the standard normal 
start1=Sys.time()
var(v) #sample variance of v using builtin function
end1=Sys.time()
tm1=end1-start1 
tm1 #elapsed time

start2=Sys.time()
myvar(v) #sample variance of v using user defined function
end2=Sys.time()
tm2=end2-start2 
tm2 #elapsed time
```

It is important to be able to write user defined functions in order to
make extensions to the language.  However, the built in functions,
which use vectorized arithmetic, with calls to more efficient languages, are typically much faster, and so the recommended choice
with even moderate sized data sets.  




##Scope:  the *scope* of a variable tells us which version of the variable is being used.
Variables can be local or global.  A variable defined within a function is local to that
function.


TRICK to remember: Variables go in but do not go out.

By this we mean that:

1. variables that are defined in the main R program keep their values (if not redefined) inside functions (variables go in)

2. variables only defined in a function have a scope limited to the function, and disappear (are undefined) back in the main program (variables do not go out)

The following exercises are just ways to familiarize you with variable scopes. You can create your own variants.

For example, test the following:

- define a and b in the main program

- modify a and b inside a function (not passing a or b as arguments)

- print a and b in the main program

- what are your conclusions?






Example: only z exists in the global environment.  *a* and *b* are defined
within the function *test*.  They are local to *test* and not available in
the global environment after the function is run.
```{r}
rm(list=ls())  #clear everything in the global environment

z=10
ls()

test=function(x){
  a=1
  b=2
  y=a+b*x
  return(y)
  }

test(z)
ls()
```

Example 2: *a*, *b* and *z* exist in the global environment.  Only *y*, and the
argument to the function, *x*, are available within the function.  When the function
*test* is defined, the values of *a* and *b* available at that time are used in
the definition.

```{r}
rm(list=ls())  #clear everything in the global environment

a=1
b=2
z=10
ls()

test=function(x){
  y=a+b*x
  print(ls()) #ls lists the variables in the local environment
              #note there is no a or b local to the function
  print(a)    #the values printed are those from the global
  print(b)    #environment when the function was defined
  return(y)
  }

test(z)
ls()
```

Example 3: *a*, *b* and *z* exists in the global environment. *a* and *b* are also
defined within the environment which is local to the function.  The local versions
are used within the function.

```{r}
rm(list=ls())  #clear everything in the global environment

a=1
b=2
z=10
ls()

test=function(x){
  a=10
  b=20
  y=a+b*x
  print(ls()) #ls lists the variables in the local environment
  return(y)
  }

output=test(z)  #now there will be a variable *output* in the global environment.

ls()
a   #these were he original, global versions of *a* and *b*
b   #the local versions were local to the function test only
```





- A more advanced example:  suppose we have one function defined within another, and the same variable name used in each function, and globally.  Before evaluating the following code, see if you can understand what the final result *f(10)+a* will
be, and also, any intermediate outputs.  What would the result be if you remove the line *a=3* in function g?  Which
value of *a* will function *g* use?  In this case, the value of *a* used in function *g* is that value which was
present in the in the environment where *g* was defined, namely *a=2*.


```{r eval=F}
rm(list=ls())
a=1
z=10

f=function(x){
  a=2
  print(c(x,a))

  g=function(x){
  a=3
  print(c(x,a))
  resultg=a+x
  print(paste("function g returns ",resultg))
  return(resultg)}

    resultf=g(x+5)+a
    print(paste("function f returns ",resultf))
    
    return(resultf)
    }

f(10)+a
```


## Advanced topic: the argument *...*

   Usually a function is called with a fixed number of arguments.  In more advanced applications, you may
   want to pass in arguments only on some occasions, and pass them directly through to other functions within
   the function you have written.    The  argument *...* is a placeholder for any number of named
   argments, which can be passed through to other functions.

   For example, suppose that we want a function which makes plots of both
  *y* vs *x*, and also a histogram of *x*.  We would like to use the function with variables of any name, but
  we would like to be able to pass in labels to the function to better identify the variables.  Let's work
  with the *Auto* data, and make plots using  *mpg* as the y-variable ,  both vs both *displacement* and *log(displacement)*
  as the x-variable, and let's pass the labels through to the plot and histogram functions.

```{r}
data=mydata=read.csv("http://faculty.marshall.usc.edu/gareth-james/ISL/Auto.csv")
attach(data)
par(mfrow=c(2,2))  #set graphics region to have two rows, two columns

plot2=function(x,y,nclassin=20,...){
 #this function first makes a plot of y vs x.  Notice that ... is passed through 
 #from the plot2 function definition
 plot(x,y,...)
 #next make a histogram of x.  Notice that the default number of histogram bars is 20
 hist(x,nclass=nclassin,...)
  }

plot2(displacement, mpg, xlab="displacement", ylab="miles per gallon")
plot2(log(displacement), mpg, xlab="log(displacement)", ylab="miles per gallon",main="mpg vs log(displacement)",nclass=40)
```



  

## Recursion - a recursive function calls itself.

- Example 1: for non-negative integer $n$, "n factorial", written as $n!$ is
defined as

$$!n = n !(n-1) = n (n-1) !(n-2) = \ldots = n(n-1)(n-2) \ldots (2)(1)$$

with the consistency condition $0! = 1$.

The following gives R code for calculating $n!$ using both recursion, and
a calculation using a *for* loop.

```{r}
fact0=function(x){
   if(x==1){return(1)
   } else   
    {return(x*fact0(x-1))}
} 
fact0(6)
#better to force the input value to be an integer
#and to include some comment as to what the function is doing
fact1=function(x){
  #returns x! for x a positive integer
  x=as.integer(x)
   if(x<0){
   print("Error: x must be a positive integer")
   return(NULL)} else {
   if(x==0|x==1){ 
   return(1)} else {
   return(x*fact1(x-1))}
   }}

fact1(6)

#fact2 calculates factorial using a for loop
fact2=function(x){
#returns x! for x a positive integer
  x=as.integer(x)
   if(x<0){
   print("Error: x must be a positive integer")
   return(NULL)} else {
   if(x==0|x==1){ 
   return(1)} else {
  temp=1
  for (i in 2:x) temp=temp*i
  return(temp)
  }}}
fact2(6)

start=Sys.time()
fact1(60) 
end=Sys.time()
end-start 

start=Sys.time()
fact2(60) 
end=Sys.time()
end-start

start=Sys.time()
factorial(60) 
end=Sys.time()
end-start
```



#Some Useful Functions for Statistics

## Arithmetic
- Arithmetic Operators


- Mathematic Functions



## Vectorized Arithmetic
```{r}
testVect = c(1,3,5,2,9,10,7,8,6)

min(testVect) # minimum
max(testVect) # maximum
mean(testVect) # mean
median(testVect) # median
quantile(testVect) # quantile
var(testVect) #variance
sd(testVect) # standard deviation

vect1 = cars$speed
vect2 = cars$dist

cov(vect1,vect2) # covariance
cor(vect1,vect2) # correlation coefficient


```

###  apply, inner product, outer product


The 'apply' function gives you a way to perform flexible operations on arrays.

The syntax is:

apply(X, MARGIN, FUN, ...)


For details on the syntax of the 'apply' family of functions, please go to:

https://www.datacamp.com/community/tutorials/r-tutorial-apply-family


```{r}
# Construct a 5x6 matrix
X <- matrix(rnorm(30), nrow=5, ncol=6)

# Sum the values of each column with `apply()`
apply(X, 2, sum)
```


```{r}
product <- outer(0:1, 0:1, "*")
#     [,1] [,2]
#[1,]    0    0
#[2,]    0    1
product2 <- outer(product,product,"*")
product
product2
```


##Probability Distributions

- key words
    - *d* : density (returns the height of the pdf)
    - *p* : distribution function (returns the cdf)
    - *q* : quantile function (returns the inverse cdf)
    - *r* : random generation

- distributions
    - *binom* : Binomial Distribution
    - *pois*  : Poisson Distribution
    - *unif* :  Uniform Distribution
    - *exp* : Exponential Distribution
    - *norm* : Normal Distribution
    - *chisq* : Chi-Squared Distribution
    - *t*   : t Distribution
    - *f*  : F Distribution




What is the probability of four or less questions answered correctly by random in a twelve question multiple choice quiz?

What is the threshold q that I should place on a random variable z drawn from a standard normal distribution to make sure that z is less than q with probability 0.75?

How can I produce a series of independent Poisson distributed counts?

Mathematical expressions of densities (pdfs):

$d \sim e^{-x^2}$

$d = {n \choose x} p^x (1-p)^{n-x}$

https://www.calvin.edu/~rpruim/courses/s341/S17/from-class/MathinRmd.html


- Examples
```{r}
# binomial  probability of having 2 successes in 10 Benoulli draws of probability 0.2
dbinom(2, size=10, prob=0.2)


dbinom(0, size=10, prob=0.2) + dbinom(1, size=10, prob=0.2) + dbinom(2, size=10, prob=0.2)

pbinom(2,size=10,prob=0.2)

runif(6,min=1,max=2)

qt(c(.025, .975), df=4)

qf(.95, df1=3, df2=4)

normSamples = rnorm(1000,mean=5,sd = 3)
hist(normSamples)

```

- *set.seed()* reproduce the results even using random
```{r}
set.seed(100)
rnorm(5)
rnorm(5)

set.seed(100) # reproduce the results
rnorm(5)
rnorm(5)
```


<!-- ## some packages (for fun) -->

<!-- - *sqldf*  -->
<!-- ```{r} -->
<!-- library(sqldf) -->
<!-- library(ISLR) -->

<!-- sqlCars = sqldf(" select name,origin,year,cylinders,horsepower -->
<!--                   from Auto  -->
<!--                   where cylinders = 8 and horsepower > 200 -->
<!--                   order by horsepower -->
<!--                 ") -->

<!-- sqlCars -->
<!-- ``` -->

<!-- - *quantmod* -->
<!-- ```{r} -->
<!-- library(quantmod) -->
<!-- getSymbols("AAPL",src="yahoo",from = Sys.Date()-500) -->
<!-- tail(AAPL) -->
<!-- barChart(AAPL) -->
<!-- getQuote("AAPL",src="yahoo") -->
<!-- ``` -->


